---
title: "ekioplot Visual Tests"
subtitle: "Palette and Theme Gallery"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(ekioplot)
library(ggplot2)
library(dplyr)
```

```{r}
#| label: helper
#| include: false

# Horizontal swatch strip, used for the group overviews below.
swatch_strip <- function(names, n = NULL) {
  par(mfrow = c(length(names), 1), mar = c(1, 8, 1, 1))
  for (pal in names) {
    cols <- as.character(ekio_pal(pal, n = n))
    barplot(rep(1, length(cols)), col = cols, border = NA, axes = FALSE,
            names.arg = rep("", length(cols)), horiz = TRUE)
    mtext(pal, side = 2, las = 1, line = 1, cex = 0.9)
  }
}
```

# Color Palettes

All palettes are reached through `ekio_pal()`. Printing one displays a
labeled swatch.

## Categorical Palettes

```{r}
#| label: categorical-palettes
#| fig-height: 6
#| fig-width: 10

swatch_strip(list_ekio_palettes("categorical"))
```

### Individual Palette Display

```{r}
#| label: show-full
#| fig-height: 2
ekio_pal("full")
```

```{r}
#| label: show-full-muted
#| fig-height: 2
ekio_pal("full_muted")
```

```{r}
#| label: show-cool3
#| fig-height: 2
ekio_pal("cool3")
```

```{r}
#| label: show-cool4
#| fig-height: 2
ekio_pal("cool4")
```

## Accent Palettes

`gold` is a fixed three-color named accent palette. `accent_blue` and
`accent_orange` return four colors by default and accept between two and six.

```{r}
#| label: accent-palettes
#| fig-height: 3
#| fig-width: 10

swatch_strip(list_ekio_palettes("accent"))
```

```{r}
#| label: show-variable-accent
#| fig-height: 2

ekio_pal("accent_blue", n = 6)
```

## EKIO Brand Palette

`ekio_brand` contains the Hokusai-derived EKIO identity colors plus white and
black. It is intended for brand assets rather than general data visualization.

```{r}
#| label: brand-palettes
#| fig-height: 2
#| fig-width: 10

swatch_strip(list_ekio_palettes("brand"))
```

## Sequential Palettes

The seven brand scales, each nine steps named `100`–`900`, light to dark.
Position and shade are aligned: `ekio_pal("blue")[7]` is `"700"`.

```{r}
#| label: sequential-palettes
#| fig-height: 6
#| fig-width: 10

swatch_strip(list_ekio_palettes("sequential"))
```

```{r}
#| label: show-seq-blue
#| fig-height: 2
ekio_pal("blue")
```

```{r}
#| label: show-seq-teal
#| fig-height: 2
ekio_pal("teal")
```

Asking for fewer colors interpolates across the whole ramp rather than
taking the lightest `n`:

```{r}
#| label: show-seq-interpolated
#| fig-height: 2
ekio_pal("blue", n = 4)
```

## Diverging Palettes

Each diverging palette's neutral pivot is its lightest color, so the visual
center lands on the data's zero.

```{r}
#| label: diverging-palettes
#| fig-height: 3
#| fig-width: 10

swatch_strip(list_ekio_palettes("diverging"))
```

```{r}
#| label: show-div-blue-orange
#| fig-height: 2
ekio_pal("blue_orange")
```

## Scientific Palettes

Colorblind-safe and perceptually uniform palettes for accessibility.

```{r}
#| label: scientific-palettes
#| fig-height: 4
#| fig-width: 10

swatch_strip(list_ekio_palettes("scientific"), n = 9)
```

# Scale Functions in Action

## Discrete Scales

```{r}
#| label: discrete-scatter
#| fig-height: 4
#| fig-width: 8

p_base <- ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  labs(title = "Discrete Color Scale", color = "Cylinders") +
  theme_ekio()

p_base + scale_color_ekio_d("cool3")
```

```{r}
#| label: discrete-bar
#| fig-height: 4
#| fig-width: 8

ggplot(mtcars, aes(factor(cyl), fill = factor(cyl))) +
  geom_bar() +
  labs(
    title = "Discrete Fill Scale: full_muted",
    x = "Cylinders",
    fill = "Cylinders"
  ) +
  scale_fill_ekio_d("full_muted") +
  theme_ekio()
```

## Continuous Scales

```{r}
#| label: continuous-scatter
#| fig-height: 4
#| fig-width: 8

ggplot(mtcars, aes(wt, mpg, color = hp)) +
  geom_point(size = 4) +
  labs(title = "Continuous Color Scale: blue", color = "Horsepower") +
  scale_color_ekio_c("blue") +
  theme_ekio()
```

```{r}
#| label: continuous-teal
#| fig-height: 4
#| fig-width: 8

ggplot(mtcars, aes(wt, mpg, color = hp)) +
  geom_point(size = 4) +
  labs(title = "Continuous Color Scale: teal", color = "Horsepower") +
  scale_color_ekio_c("teal") +
  theme_ekio()
```

## Diverging Scales

Diverging palettes go through the continuous scales, same as sequential.

```{r}
#| label: diverging-scatter
#| fig-height: 4
#| fig-width: 8

# Create centered data
df <- mtcars |>
  mutate(mpg_centered = mpg - mean(mpg))

ggplot(df, aes(wt, mpg_centered, color = mpg_centered)) +
  geom_point(size = 4) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "Diverging Color Scale: blue_orange",
    color = "MPG (centered)",
    y = "MPG (deviation from mean)"
  ) +
  scale_color_ekio_c("blue_orange") +
  theme_ekio()
```

```{r}
#| label: diverging-heatmap
#| fig-height: 5
#| fig-width: 8

# Correlation heatmap
cor_mat <- cor(mtcars[, c("mpg", "disp", "hp", "wt", "qsec")])
cor_df <- as.data.frame(as.table(cor_mat))
names(cor_df) <- c("Var1", "Var2", "Correlation")

ggplot(cor_df, aes(Var1, Var2, fill = Correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Correlation, 2)), size = 3, color = "white") +
  labs(title = "Correlation Heatmap with Diverging Scale") +
  scale_fill_ekio_c("blue_red") +
  theme_ekio(grid = "none") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Theme Variations

## Grid Options

```{r}
#| label: theme-grid-y
#| fig-height: 3.5
#| fig-width: 6

p <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point(color = ekio_pal("blue")["700"], size = 3) +
  labs(title = "grid = \"y\" (default)")

p + theme_ekio(grid = "y")
```

```{r}
#| label: theme-grid-x
#| fig-height: 3.5
#| fig-width: 6

p + labs(title = "grid = \"x\"") + theme_ekio(grid = "x")
```

```{r}
#| label: theme-grid-xy
#| fig-height: 3.5
#| fig-width: 6

p + labs(title = "grid = \"xy\"") + theme_ekio(grid = "xy")
```

```{r}
#| label: theme-grid-none
#| fig-height: 3.5
#| fig-width: 6

p + labs(title = "grid = \"none\"") + theme_ekio(grid = "none")
```

## Base Size

`base_size` scales the whole type hierarchy — use a larger value for slides.

```{r}
#| label: theme-base-size
#| fig-height: 4
#| fig-width: 8

ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) +
  geom_point(size = 4) +
  scale_color_ekio_d("cool3") +
  labs(
    title = "Larger Base Size",
    subtitle = "base_size = 16 for presentation contexts",
    color = "Cylinders"
  ) +
  theme_ekio(base_size = 16)
```

# Recipe Functions

## Histogram with Aesthetic Detection

```{r}
#| label: recipe-histogram-static
#| fig-height: 4
#| fig-width: 7

# Static color
ekio_histogram(mtcars, x = mpg, fill = "steelblue", title = "Static fill color")
```

```{r}
#| label: recipe-histogram-variable
#| fig-height: 4
#| fig-width: 7

# Variable mapping
ekio_histogram(mtcars, x = mpg, fill = factor(cyl), title = "Fill by variable")
```

## Bar Plot

```{r}
#| label: recipe-barplot
#| fig-height: 4
#| fig-width: 7

# ekio_barplot needs an explicit y; summarize first
mpg_by_cyl <- mtcars |>
  summarise(mean_mpg = mean(mpg), .by = cyl) |>
  mutate(cyl = factor(cyl))

ekio_barplot(
  mpg_by_cyl,
  x = cyl,
  y = mean_mpg,
  fill = cyl,
  title = "Bar plot with EKIO palette"
)
```

## Scatter Plot

```{r}
#| label: recipe-scatter
#| fig-height: 4
#| fig-width: 7

ekio_scatterplot(
  mtcars,
  x = wt,
  y = mpg,
  color = factor(cyl),
  title = "Scatter plot by group"
)
```

## Line Plot

```{r}
#| label: recipe-line
#| fig-height: 4
#| fig-width: 7

# Create time series-like data
df_line <- data.frame(
  time = rep(1:10, 3),
  value = c(cumsum(rnorm(10)), cumsum(rnorm(10)), cumsum(rnorm(10))),
  group = rep(c("A", "B", "C"), each = 10)
)

ekio_lineplot(
  df_line,
  x = time,
  y = value,
  color = group,
  title = "Line plot by group"
)
```

# Brand Scale Access

Single colors come off the brand scales by shade. `as.character()` strips the
swatch class when only hex codes are wanted.

```{r}
#| label: scale-access

as.character(ekio_pal("blue"))

# Position and shade agree
ekio_pal("blue")["700"]
ekio_pal("blue")[7]

# The full categorical palette
as.character(ekio_pal("full"))
```

# Summary

This document showcases the visual elements of ekioplot:

- **Categorical palettes**: `r length(list_ekio_palettes("categorical"))` palettes for categorical data
- **Brand palette**: `ekio_brand` for EKIO identity assets
- **Accent palettes**: `r length(list_ekio_palettes("accent"))` named and variable-length accent sets
- **Sequential palettes**: `r length(list_ekio_palettes("sequential"))` brand scales for continuous data
- **Diverging palettes**: `r length(list_ekio_palettes("diverging"))` palettes for centered data
- **Scientific palettes**: `r length(list_ekio_palettes("scientific"))` colorblind-safe palettes for accessibility
- **Theme variations**: `theme_ekio()` with grid and base size options
- **Recipe functions**: Smart aesthetic detection for intuitive plotting
