---
title: "ekioplot v2.0 Visual Tests"
subtitle: "Palette and Theme Gallery"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(ekioplot)
library(ggplot2)
library(dplyr)
```

# Color Palettes

## EKIO Qualitative Palettes

```{r}
#| label: qualitative-palettes
#| fig-height: 6
#| fig-width: 10

palettes <- list_ekio_palettes("qualitative")

par(mfrow = c(length(palettes), 1), mar = c(1, 8, 1, 1))
for (pal in palettes) {
  cols <- ekio_pal(pal)
  barplot(rep(1, length(cols)), col = cols, border = NA, axes = FALSE,
          names.arg = rep("", length(cols)), horiz = TRUE)
  mtext(pal, side = 2, las = 1, line = 1, cex = 0.9)
}
```

### Individual Palette Display

```{r}
#| label: show-contrast
#| fig-height: 2
show_ekio_palette("contrast")
```

```{r}
#| label: show-cool
#| fig-height: 2
show_ekio_palette("cool")
```

```{r}
#| label: show-full
#| fig-height: 2
show_ekio_palette("full")
```

```{r}
#| label: show-muted
#| fig-height: 2
show_ekio_palette("muted")
```

## EKIO Sequential Palettes

```{r}
#| label: sequential-palettes
#| fig-height: 4
#| fig-width: 10

seq_palettes <- list_ekio_palettes("sequential")

par(mfrow = c(length(seq_palettes), 1), mar = c(1, 8, 1, 1))
for (pal in seq_palettes) {
  cols <- ekio_seq_pal(pal, n = 9)
  barplot(rep(1, length(cols)), col = cols, border = NA, axes = FALSE,
          names.arg = rep("", length(cols)), horiz = TRUE)
  mtext(pal, side = 2, las = 1, line = 1, cex = 0.9)
}
```

```{r}
#| label: show-seq-blue
#| fig-height: 2
show_ekio_palette("blue", type = "sequential")
```

```{r}
#| label: show-seq-teal
#| fig-height: 2
show_ekio_palette("teal", type = "sequential")
```

## EKIO Diverging Palettes

```{r}
#| label: diverging-palettes
#| fig-height: 3
#| fig-width: 10

div_palettes <- list_ekio_palettes("diverging")

par(mfrow = c(length(div_palettes), 1), mar = c(1, 8, 1, 1))
for (pal in div_palettes) {
  cols <- ekio_div_pal(pal, n = 9)
  barplot(rep(1, length(cols)), col = cols, border = NA, axes = FALSE,
          names.arg = rep("", length(cols)), horiz = TRUE)
  mtext(pal, side = 2, las = 1, line = 1, cex = 0.9)
}
```

```{r}
#| label: show-div-blue-orange
#| fig-height: 2
show_ekio_palette("blue_orange", type = "diverging")
```

## External Palettes (Non-EKIO)

```{r}
#| label: external-palettes
#| fig-height: 4
#| fig-width: 10

ext_palettes <- names(list_external_palettes())

par(mfrow = c(length(ext_palettes), 1), mar = c(1, 8, 1, 1))
for (pal in ext_palettes) {
  cols <- external_pal(pal, n = 9)
  barplot(rep(1, length(cols)), col = cols, border = NA, axes = FALSE,
          names.arg = rep("", length(cols)), horiz = TRUE)
  mtext(pal, side = 2, las = 1, line = 1, cex = 0.9)
}
```

# Scale Functions in Action

## Discrete Scales

```{r}
#| label: discrete-scatter
#| fig-height: 4
#| fig-width: 8

p_base <- ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  labs(title = "Discrete Color Scale", color = "Cylinders") +
  theme_ekio()

p_base + scale_color_ekio_d("contrast")
```

```{r}
#| label: discrete-bar
#| fig-height: 4
#| fig-width: 8

ggplot(mtcars, aes(factor(cyl), fill = factor(cyl))) +
  geom_bar() +
  labs(title = "Discrete Fill Scale: cool", x = "Cylinders", fill = "Cylinders") +
  scale_fill_ekio_d("cool") +
  theme_ekio()
```

## Continuous Scales

```{r}
#| label: continuous-scatter
#| fig-height: 4
#| fig-width: 8

ggplot(mtcars, aes(wt, mpg, color = hp)) +
  geom_point(size = 4) +
  labs(title = "Continuous Color Scale: blue", color = "Horsepower") +
  scale_color_ekio_c("blue") +
  theme_ekio()
```

```{r}
#| label: continuous-teal
#| fig-height: 4
#| fig-width: 8

ggplot(mtcars, aes(wt, mpg, color = hp)) +
  geom_point(size = 4) +
  labs(title = "Continuous Color Scale: teal", color = "Horsepower") +
  scale_color_ekio_c("teal") +
  theme_ekio()
```

## Diverging Scales

```{r}
#| label: diverging-scatter
#| fig-height: 4
#| fig-width: 8

# Create centered data
df <- mtcars |>
  mutate(mpg_centered = mpg - mean(mpg))

ggplot(df, aes(wt, mpg_centered, color = mpg_centered)) +
  geom_point(size = 4) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "Diverging Color Scale: blue_orange",
    color = "MPG (centered)",
    y = "MPG (deviation from mean)"
  ) +
  scale_color_ekio_div("blue_orange") +
  theme_ekio()
```

```{r}
#| label: diverging-heatmap
#| fig-height: 5
#| fig-width: 8

# Correlation heatmap
cor_mat <- cor(mtcars[, c("mpg", "disp", "hp", "wt", "qsec")])
cor_df <- as.data.frame(as.table(cor_mat))
names(cor_df) <- c("Var1", "Var2", "Correlation")

ggplot(cor_df, aes(Var1, Var2, fill = Correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Correlation, 2)), size = 3, color = "white") +
  labs(title = "Correlation Heatmap with Diverging Scale") +
  scale_fill_ekio_div("blue_red") +
  theme_ekio(grid = "none") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Theme Variations

## Grid Options

```{r}
#| label: theme-grid-y
#| fig-height: 3.5
#| fig-width: 6

p <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point(color = ekio_accent["blue"], size = 3) +
  labs(title = "grid = \"y\" (default)")

p + theme_ekio(grid = "y")
```

```{r}
#| label: theme-grid-x
#| fig-height: 3.5
#| fig-width: 6

p + labs(title = "grid = \"x\"") + theme_ekio(grid = "x")
```

```{r}
#| label: theme-grid-xy
#| fig-height: 3.5
#| fig-width: 6

p + labs(title = "grid = \"xy\"") + theme_ekio(grid = "xy")
```

```{r}
#| label: theme-grid-none
#| fig-height: 3.5
#| fig-width: 6

p + labs(title = "grid = \"none\"") + theme_ekio(grid = "none")
```

## Theme Variants

```{r}
#| label: theme-presentation
#| fig-height: 4
#| fig-width: 8

ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) +
  geom_point(size = 4) +
  scale_color_ekio_d("contrast") +
  labs(
    title = "Presentation Theme",
    subtitle = "Larger text and margins for slides",
    color = "Cylinders"
  ) +
  theme_ekio_presentation()
```

```{r}
#| label: theme-map
#| fig-height: 4
#| fig-width: 6

# Simulated map-like plot
ggplot(mtcars, aes(wt, mpg, color = hp)) +
  geom_point(size = 4) +
  scale_color_ekio_c("blue") +
  labs(title = "Map Theme (no axes)", color = "HP") +
  theme_ekio_map()
```

# Recipe Functions

## Histogram with Aesthetic Detection

```{r}
#| label: recipe-histogram-static
#| fig-height: 4
#| fig-width: 7

# Static color
ekio_histogram(mtcars, x = mpg, fill = "steelblue", title = "Static fill color")
```

```{r}
#| label: recipe-histogram-variable
#| fig-height: 4
#| fig-width: 7

# Variable mapping
ekio_histogram(mtcars, x = mpg, fill = factor(cyl), title = "Fill by variable")
```

## Bar Plot

```{r}
#| label: recipe-barplot
#| fig-height: 4
#| fig-width: 7

ekio_barplot(mtcars, x = factor(cyl), fill = factor(cyl), title = "Bar plot with EKIO palette")
```

## Scatter Plot

```{r}
#| label: recipe-scatter
#| fig-height: 4
#| fig-width: 7

ekio_scatterplot(mtcars, x = wt, y = mpg, color = factor(cyl), title = "Scatter plot by group")
```

## Line Plot

```{r}
#| label: recipe-line
#| fig-height: 4
#| fig-width: 7

# Create time series-like data
df_line <- data.frame(
  time = rep(1:10, 3),
  value = c(cumsum(rnorm(10)), cumsum(rnorm(10)), cumsum(rnorm(10))),
  group = rep(c("A", "B", "C"), each = 10)
)

ekio_lineplot(df_line, x = time, y = value, color = group, title = "Line plot by group")
```

# Color Scale Vectors

```{r}
#| label: color-vectors

# Display ekio_blue
cat("ekio_blue:\n")
print(ekio_blue)

# Display ekio_accent
cat("\nekio_accent:\n")
print(ekio_accent)
```

# Summary

This document showcases all visual elements of ekioplot v2.0:

- **Qualitative palettes**: `r length(list_ekio_palettes("qualitative"))` palettes for categorical data
- **Sequential palettes**: `r length(list_ekio_palettes("sequential"))` palettes for continuous data
- **Diverging palettes**: `r length(list_ekio_palettes("diverging"))` palettes for centered data
- **External palettes**: `r length(list_external_palettes())` non-EKIO palettes for accessibility
- **Theme variations**: `theme_ekio()` with grid options, plus `theme_ekio_presentation()` and `theme_ekio_map()`
- **Recipe functions**: Smart aesthetic detection for intuitive plotting
