---
title: "Getting Started with ekioplot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ekioplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  dpi = 150
)
```

```{r setup}
library(ekioplot)
library(ggplot2)
```

## EKIO Theme

`theme_ekio()` applies EKIO's visual identity to any ggplot2 plot. It builds
on `theme_minimal()` with curated typography, spacing, and color choices.

```{r theme-basic}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point(color = ekio_blue["700"], size = 2.5) +
  labs(
    title = "Fuel Efficiency vs. Weight",
    subtitle = "Motor Trend Car Road Tests (1974)",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon"
  ) +
  theme_ekio()
```

The `grid` parameter controls which major grid lines are drawn:

```{r theme-grid}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point(color = ekio_blue["700"]) +
  theme_ekio(grid = "xy")
```

Use `theme_ekio_map()` for spatial visualizations — it removes axes and
repositions the legend.

## Color Palettes

ekioplot ships with ~30 palettes across five categories. Use
`list_ekio_palettes()` to explore them:

```{r list-palettes}
str(list_ekio_palettes())
```

Access any palette with `ekio_pal()`:

```{r palette-access}
ekio_pal("contrast")
ekio_pal("blue", n = 5)
```

### Palette types

- **Categorical**: `contrast`, `cool`, `minimal`, `full`, `muted`, `binary`,
  `political`
- **Small-group**: `duo_warm`, `duo_cool`, `trio_bold`, `trio_cool`,
  `quad_earth`, `quad_vivid`
- **Scientific**: `okabe_ito`, `viridis`, `inferno`, `plasma`
- **Sequential**: `blue`, `teal`, `gray`, `orange`, `purple`, `red`, `green`,
  `amber`
- **Diverging**: `blue_orange`, `blue_red`, `teal_orange`

Visualize any palette with `show_ekio_palette()`:

```{r show-palette}
show_ekio_palette("contrast")
```

## Scale Functions

ekioplot provides ggplot2 scales for both discrete and continuous data.

### Discrete scales

```{r scale-discrete}
ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  scale_color_ekio_d("contrast") +
  labs(color = "Cylinders") +
  theme_ekio(grid = "xy")
```

### Continuous scales

Sequential and diverging palettes work with continuous data:

```{r scale-continuous}
ggplot(mtcars, aes(wt, mpg, color = hp)) +
  geom_point(size = 3) +
  scale_color_ekio_c("blue") +
  labs(color = "Horsepower") +
  theme_ekio(grid = "xy")
```

Fill variants are available as `scale_fill_ekio_d()` and
`scale_fill_ekio_c()`.

## Recipe Functions

Recipe functions are high-level wrappers that create complete,
publication-ready plots with smart defaults.

### Histogram

```{r recipe-histogram}
ekio_histogram(mtcars, mpg)
```

### Bar plot

```{r recipe-barplot}
cyl_counts <- as.data.frame(table(cyl = mtcars$cyl))
names(cyl_counts)[2] <- "n"
ekio_barplot(cyl_counts, cyl, n)
```

### Scatter plot

```{r recipe-scatter}
ekio_scatterplot(mtcars, wt, mpg, color = factor(cyl))
```

### Area plot

```{r recipe-area}
data(fuels)
world_fuels <- fuels[fuels$entity == "World" & fuels$year >= 1950, ]
ekio_areaplot(world_fuels, year, consumption_gwh, fill = fuel)
```

### Smart aesthetic detection

Recipe functions automatically detect whether the color/fill argument is:

- **Missing** — uses EKIO blue as default
- **A color string** (e.g., `"steelblue"`) — uses that color directly
- **A variable** — maps it and applies the appropriate EKIO scale

```{r recipe-static-color}
ekio_histogram(mtcars, mpg, fill = "coral")
```

## Color Scales

Four named color scales are exported for direct use: `ekio_blue`, `ekio_gray`,
`ekio_teal`, and `ekio_orange`. Each provides 10 shades from `"50"` (lightest)
to `"900"` (darkest).

```{r color-vectors}
ekio_blue["700"]
ekio_gray["300"]
```

Named accent colors are available in `ekio_accent`:

```{r accent-colors}
ekio_accent
```

## GT Tables

Apply EKIO styling to gt tables with `gt_theme_ekio()`:

```{r gt-theme, eval = requireNamespace("gt", quietly = TRUE)}
library(gt)

head(mtcars[, 1:5], 8) |>
  gt() |>
  gt_theme_ekio(add_footer = FALSE)
```
