---
title: "Getting Started with ekioplot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ekioplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(ekioplot)
library(ggplot2)
```

## EKIO Theme

`theme_ekio()` applies EKIO's visual identity to any ggplot2 plot. It builds
on `theme_minimal()` with curated typography, spacing, and color choices.

```{r theme-basic}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point(color = ekio_pal("blue")["700"], size = 2.5) +
  labs(
    title = "Fuel Efficiency vs. Weight",
    subtitle = "Motor Trend Car Road Tests (1974)",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon"
  ) +
  theme_ekio()
```

The `grid` parameter controls which major grid lines are drawn:

```{r theme-grid}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point(color = ekio_pal("blue")["700"]) +
  theme_ekio(grid = "xy")
```

## Color Palettes

ekioplot ships with ~30 palettes across five categories. Use
`list_ekio_palettes()` to explore them:

```{r list-palettes}
str(list_ekio_palettes())
```

Access any palette with `ekio_pal()`:

```{r palette-access}
ekio_pal("contrast")
ekio_pal("blue", n = 5)
```

### Palette types

- **Categorical**: `contrast`, `cool`, `minimal`, `full`, `muted`,
  `muted_warm`, `binary`, `political`
- **Highlight**: `highlight_blue`, `highlight_orange`, `highlight_teal`,
  `highlight_red`
- **Small-group**: `duo_warm`, `duo_cool`, `trio_bold`, `trio_cool`,
  `quad_earth`, `quad_vivid`
- **Scientific**: `okabe_ito`, `viridis`, `inferno`, `plasma`
- **Sequential**: `blue`, `gray`, `stone`, `teal`, `green`, `orange`, `red`
- **Accent**: `gold`
- **Diverging**: `blue_orange`, `blue_red`, `teal_orange`

`ekio_pal()` displays a swatch when printed:

```{r show-palette}
ekio_pal("contrast")
```

## Scale Functions

ekioplot provides ggplot2 scales for both discrete and continuous data.

### Discrete scales

```{r scale-discrete}
ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  scale_color_ekio_d("contrast") +
  labs(color = "Cylinders") +
  theme_ekio(grid = "xy")
```

### Continuous scales

Sequential and diverging palettes work with continuous data:

```{r scale-continuous}
ggplot(mtcars, aes(wt, mpg, color = hp)) +
  geom_point(size = 3) +
  scale_color_ekio_c("blue") +
  labs(color = "Horsepower") +
  theme_ekio(grid = "xy")
```

Fill variants are available as `scale_fill_ekio_d()` and
`scale_fill_ekio_c()`.

## Recipe Functions

Recipe functions are high-level wrappers that create complete,
publication-ready plots with smart defaults.

### Histogram

```{r recipe-histogram}
ekio_histogram(mtcars, mpg)
```

### Bar plot

```{r recipe-barplot}
cyl_counts <- as.data.frame(table(cyl = mtcars$cyl))
names(cyl_counts)[2] <- "n"
ekio_barplot(cyl_counts, cyl, n)
```

### Scatter plot

```{r recipe-scatter}
ekio_scatterplot(mtcars, wt, mpg, color = factor(cyl))
```

### Area plot

```{r recipe-area}
data(fuels)
world_fuels <- fuels[fuels$entity == "World" & fuels$year >= 1950, ]
ekio_areaplot(world_fuels, year, consumption_gwh, fill = fuel)
```

### Smart aesthetic detection

Recipe functions automatically detect whether the color/fill argument is:

- **Missing** — uses EKIO blue as default
- **A color string** (e.g., `"steelblue"`) — uses that color directly
- **A variable** — maps it and applies the appropriate EKIO scale

```{r recipe-static-color}
ekio_histogram(mtcars, mpg, fill = "coral")
```

## Brand Scales

Every brand color is reached through `ekio_pal()`. The seven brand scales —
`"blue"`, `"gray"`, `"stone"`, `"teal"`, `"green"`, `"orange"`, and `"red"` —
are nine-step ramps running light to dark, named by shade. All seven sit on one
lightness spine, so a given shade carries the same visual weight in every
family:

```{r brand-scale}
ekio_pal("gray")
```

Position and shade are aligned by construction, so element `i` is always
shade `i * 100`. Index whichever way reads better:

```{r brand-shades}
ekio_pal("blue")["700"]
ekio_pal("blue")[7]
```

Because these are the same objects used for continuous fills, asking for
fewer colors interpolates across the whole ramp rather than returning the
lightest few:

```{r brand-interpolate}
ekio_pal("blue", n = 3)
```

## Tables

gt table styling lives in a companion package,
[ekiotable](https://github.com/viniciusoike/ekiotable). It reads brand tokens
from ekioplot via `ekio_pal()`, so tables and charts share one palette
without either package duplicating color definitions.
