---
title: "Color accessibility"
description: >
  Which EKIO colors work with black text, which need white text, and how
  to check contrast against WCAG guidelines.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4.5,
  dpi = 150
)
```

```{r setup, message = FALSE}
library(ekioplot)
library(ggplot2)
```

Text placed on a colored background — bar labels, gt table cells, annotation
boxes — needs enough contrast to stay readable. The [WCAG
2.1](https://www.w3.org/TR/WCAG21/#contrast-minimum) guidelines quantify this
as a **contrast ratio** between text and background, from 1 (identical
colors) to 21 (black on white):

| Level | Normal text | Large text (≥ 18pt) |
|-------|-------------|---------------------|
| AA    | ≥ 4.5       | ≥ 3.0               |
| AAA   | ≥ 7.0       | ≥ 4.5               |

`ekioplot` provides two helpers:

- `ekio_contrast()` computes the WCAG contrast ratio between two colors.
- `ekio_text_on()` picks the more readable text color (black or white) for a
  given background.

```{r}
ekio_contrast("white", ekio_pal("blue")["700"])
ekio_text_on(ekio_pal("blue")["700"])
```

## Black or white text on the EKIO scales

The chart below shows every shade of the four EKIO color scales, labeled in
the text color that `ekio_text_on()` selects, with its contrast ratio in
parentheses.

```{r shades, fig.height = 6}
scales_df <- do.call(rbind, lapply(
  c("blue", "teal", "orange", "gray"),
  function(scale_name) {
    hex <- ekio_pal(scale_name)
    data.frame(
      scale = scale_name,
      shade = factor(names(hex), levels = rev(names(hex))),
      hex = as.character(hex)
    )
  }
))

scales_df$text_color <- ekio_text_on(scales_df$hex)
scales_df$ratio <- ifelse(
  scales_df$text_color == "white",
  ekio_contrast("white", scales_df$hex),
  ekio_contrast("black", scales_df$hex)
)
scales_df$label <- sprintf(
  "%s (%.1f)", scales_df$hex, scales_df$ratio
)

ggplot(scales_df, aes(x = scale, y = shade, fill = hex)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = label, color = text_color), size = 3) +
  scale_fill_identity() +
  scale_color_identity() +
  scale_x_discrete(position = "top") +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())
```

The cutoff is *not* the same across scales. Blue flips to white text at
`500`, teal and gray at `700`, and orange only at `800` — the warmer, more
luminous hues stay readable against black text much further down the ramp.
Call `ekio_text_on()` rather than memorizing a cutoff.

## Accent colors

`ekio_pal("full")` carries one accent per brand scale, in scale order. These
are mid-to-dark tones, so most of them pair with white text — but not all
reach AA for normal-size text.

```{r accents, fig.height = 3}
accent_names <- c(
  "blue", "orange", "teal", "amber", "purple", "red", "green", "gray"
)

accents_df <- data.frame(
  name = factor(accent_names, levels = accent_names),
  hex = as.character(ekio_pal("full"))
)
accents_df$text_color <- ekio_text_on(accents_df$hex)
accents_df$ratio <- ekio_contrast(accents_df$text_color, accents_df$hex)
accents_df$label <- sprintf("%s\n%.1f", accents_df$name, accents_df$ratio)

ggplot(accents_df, aes(x = name, y = 1, fill = hex)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = label, color = text_color), size = 3.5, lineheight = 1) +
  scale_fill_identity() +
  scale_color_identity() +
  labs(x = NULL, y = NULL) +
  theme_void()
```

A full compliance table:

```{r}
accents_df$aa <- accents_df$ratio >= 4.5
accents_df$aa_large <- accents_df$ratio >= 3.0
accents_df$aaa <- accents_df$ratio >= 7.0

accents_df[, c("name", "hex", "text_color", "ratio", "aa", "aa_large", "aaa")]
```

Colors that pass `aa_large` but not `aa` (such as `orange` and `amber` with
white text) are fine for large display text — big value labels, headline
numbers — but should not carry small annotations. For small text on those
fills, use a dark text color such as `ekio_pal("gray")["900"]`:

```{r}
ekio_contrast(
  ekio_pal("gray")["900"],
  accents_df$hex[accents_df$name == "amber"]
)
```

## Using `ekio_text_on()` in plots

Pass the fill colors through `ekio_text_on()` to color labels, so they stay
readable regardless of the underlying shade:

```{r barplot-labels}
sales <- data.frame(
  region = c("North", "Northeast", "Southeast", "South"),
  value = c(42, 65, 118, 87)
)
sales$fill <- ekio_pal("blue", n = 4)

ggplot(sales, aes(x = reorder(region, value), y = value, fill = fill)) +
  geom_col() +
  geom_text(
    aes(label = value, color = ekio_text_on(fill)),
    hjust = 1.3,
    fontface = "bold"
  ) +
  scale_fill_identity() +
  scale_color_identity() +
  coord_flip() +
  labs(x = NULL, y = NULL, title = "Readable labels on any shade") +
  theme_ekio(grid = "x")
```

## Guidelines

- Default to `ekio_pal("blue")["700"]` (the primary brand blue) for filled
  elements that carry white text: it passes AAA (ratio
  `r round(ekio_contrast("white", ekio_pal("blue")["700"]), 1)`).
- Light backgrounds (`100`–`200` shades) are for panels and subtle fills;
  always use dark text on them.
- Don't rely on the black/white cutoff from memory when generating fills
  programmatically — compute it with `ekio_text_on()`.
- Contrast ratios apply to text and essential icons, not to adjacent chart
  fills; for distinguishing series, palette choice (see `ekio_pal()`)
  matters more than contrast against the background.
