---
title: "Palette Gallery"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(ekioplot)
library(reactable)
```

Every palette `ekioplot` ships, at full size. Colors are read live from the
package, so this page is always in sync with
`inst/ekio-palettes.yaml` — the source of truth for the brand.

Each chip is labelled with its hex code, drawn in black or white by
`ekio_text_on()`, the same WCAG contrast helper the palette swatches use.

The seven brand scales are generated from one OKLCH specification: a shared
lightness spine anchored on the brand navy, a shared chroma arc, and a hue
path per family. A given shade therefore carries the same visual weight in
every family. The scientific palettes are not EKIO color — they come from
matplotlib and from Okabe & Ito, and `inst/COPYRIGHTS` carries the notices.

```{r swatch-helpers}
palette_names <- unlist(list_ekio_palettes(), use.names = FALSE)

type_labels <- c(
  accent = "Accent",
  categorical = "Categorical",
  highlight = "Highlight",
  small_group = "Small group",
  sequential = "Sequential",
  diverging = "Diverging",
  scientific = "Scientific"
)

# One palette -> one labelled strip of chips.
swatch <- function(name) {
  pal <- ekio_pal(name)
  hex <- as.character(pal)
  ink <- unname(ekio_text_on(hex, dark = as.character(ekio_pal("gray")["900"])))

  chips <- paste0(
    '<div class="ekio-chip" style="background:', hex, ';color:', ink,
    '" title="', hex, '">', hex, "</div>",
    collapse = ""
  )

  paste0(
    '<div class="ekio-swatch">',
    '<div class="ekio-swatch-name"><code>', name, "</code>",
    '<span class="ekio-swatch-n">', length(hex), " colors</span></div>",
    '<div class="ekio-swatch-strip">', chips, "</div>",
    "</div>"
  )
}

gallery <- function(type) {
  strips <- vapply(list_ekio_palettes(type), swatch, character(1))
  knitr::asis_output(paste0(strips, collapse = "\n"))
}
```

## All colors

Every color in one searchable table. Click a hex to copy it; expand a palette
to see its colors, or click its **copy** button to take the whole thing as an R
vector.

```{r copy-helpers, results = "asis"}
# Palettes are handed to the page as JSON so the click handlers can look one up
# by name - that keeps quoting out of the inline onclick attributes below.
palette_json <- function() {
  entries <- vapply(
    palette_names,
    function(name) {
      hex <- as.character(ekio_pal(name))
      paste0('"', name, '":["', paste(hex, collapse = '","'), '"]')
    },
    character(1)
  )
  paste0("{", paste(entries, collapse = ","), "}")
}

knitr::asis_output(paste0(
  "<script>\n",
  "window.ekioPalettes = ", palette_json(), ";\n",
  "function ekioCopy(text, btn) {\n",
  "  if (!navigator.clipboard) return;\n",
  "  navigator.clipboard.writeText(text);\n",
  "  var label = btn.querySelector('.ekio-copy-label');\n",
  "  if (!label) return;\n",
  "  if (!btn.dataset.label) btn.dataset.label = label.textContent;\n",
  "  label.textContent = 'copied';\n",
  "  btn.classList.add('is-copied');\n",
  "  setTimeout(function () {\n",
  "    label.textContent = btn.dataset.label;\n",
  "    btn.classList.remove('is-copied');\n",
  "  }, 900);\n",
  "}\n",
  "function ekioCopyPalette(name, btn) {\n",
  "  var hex = window.ekioPalettes[name] || [];\n",
  "  var vec = 'c(' + hex.map(function (h) { return '\"' + h + '\"'; })",
  ".join(', ') + ')';\n",
  "  ekioCopy(vec, btn);\n",
  "}\n",
  "</script>"
))
```

```{r color-table}
# One row per color, across every palette
color_table <- function() {
  groups <- list_ekio_palettes()

  rows <- lapply(names(groups), function(type) {
    per_palette <- lapply(groups[[type]], function(name) {
      pal <- ekio_pal(name)
      hex <- as.character(pal)
      data.frame(
        palette = name,
        type = type_labels[[type]],
        position = seq_along(hex),
        shade = if (is.null(names(pal))) NA_character_ else names(pal),
        hex = hex
      )
    })
    do.call(rbind, per_palette)
  })

  do.call(rbind, rows)
}

copy_hex <- JS("
function(cellInfo) {
  var hex = cellInfo.value
  return '<button type=\"button\" class=\"ekio-copy\" ' +
    'onclick=\"ekioCopy(\\'' + hex + '\\', this)\" ' +
    'title=\"Copy ' + hex + '\">' +
    '<span class=\"ekio-copy-chip\" style=\"background:' + hex + '\"></span>' +
    '<span class=\"ekio-copy-label\">' + hex + '</span></button>'
}")

copy_palette <- JS("
function(cellInfo) {
  var name = cellInfo.row.palette
  return '<button type=\"button\" class=\"ekio-copy\" ' +
    'onclick=\"ekioCopyPalette(\\'' + name + '\\', this)\" ' +
    'title=\"Copy all colors as an R vector\">' +
    '<span class=\"ekio-copy-label\">copy c(...)</span></button>'
}")

reactable(
  color_table(),
  groupBy = "palette",
  searchable = TRUE,
  pagination = FALSE,
  highlight = TRUE,
  compact = TRUE,
  columns = list(
    palette = colDef(name = "Palette", minWidth = 150),
    # Shown once on the palette row rather than repeated down every color
    type = colDef(
      name = "Type",
      aggregate = "unique",
      cell = JS("function() { return '' }"),
      minWidth = 110
    ),
    # The group row already carries a count, so this only labels color rows
    position = colDef(name = "#", maxWidth = 60),
    shade = colDef(name = "Shade", na = "-", maxWidth = 80),
    hex = colDef(
      name = "Hex",
      html = TRUE,
      cell = copy_hex,
      aggregated = copy_palette,
      minWidth = 150
    )
  ),
  theme = reactableTheme(
    color = "var(--bs-body-color)",
    backgroundColor = "transparent",
    borderColor = "var(--ekio-rule)",
    highlightColor = "rgba(var(--bs-primary-rgb), 0.07)",
    cellPadding = "6px 8px",
    style = list(fontSize = "0.875rem"),
    searchInputStyle = list(width = "100%")
  )
)
```

## Accent

Named tokens rather than a ramp. Gold has no nine-step scale because dark
yellow is brown — past the middle of the spine a gold ramp stops being gold.
The three sit on the same lightness rungs as scale shades 300, 400 and 500,
and `deep` is the one that can carry type.

```{r accent, results = "asis"}
gallery("accent")
```

## Categorical

For unordered groups. Each series sits on its own rung of the lightness
spine, so `contrast` stays readable in grayscale and under deuteranopia.
Past five categories `full` separates by hue rather than lightness, so it
does not survive grayscale printing. `cool` and `minimal` stay within the
primary blues and grays; `muted` and `muted_warm` are gray-only; `binary` and
`political` are fixed two-color pairings.

```{r categorical, results = "asis"}
gallery("categorical")
```

## Highlight

One accent against three receding grays. The series carrying the argument
speaks and the rest stay legible context — order the factor so the level you
want emphasised comes first.

```{r highlight, results = "asis"}
gallery("highlight")
```

## Small group

Tuned for a known number of series — reach for these instead of truncating a
longer palette, which can leave you with two neighbouring hues.

```{r small-group, results = "asis"}
gallery("small_group")
```

## Sequential

The seven brand scales, light to dark. Position `i` is shade `i * 100`, so
`ekio_pal("blue")["700"]` is the primary blue. These work as discrete palettes
for ordered categories and as the ramp behind `scale_color_ekio_c()`.

```{r sequential, results = "asis"}
gallery("sequential")
```

## Diverging

For values with a meaningful midpoint. The pivot is a near-neutral tone,
lighter than both arms, so the visual center of the scale lands on zero.

```{r diverging, results = "asis"}
gallery("diverging")
```

## Scientific

Established palettes included for accessibility and continuity with published
work: `okabe_ito` is colorblind-safe, and the three matplotlib ramps are
perceptually uniform.

```{r scientific, results = "asis"}
gallery("scientific")
```

## Using a palette

Pass any name above to a scale function, or call `ekio_pal()` directly to pull
hex codes out.

```{r usage, echo = TRUE, eval = FALSE}
library(ggplot2)

ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  scale_color_ekio_d("contrast") +
  theme_ekio()

# Named access to any brand scale
ekio_pal("teal")["600"]
#>     600
#> "#006261"

# Interpolated to an arbitrary length
ekio_pal("blue_orange", n = 15)
```

`list_ekio_palettes()` returns these names programmatically, and
`ekio_pal()` prints a swatch when called at the console.
